Normalization by Evaluation for Call-By-Push-Value and Polarized Lambda Calculus

### 2019-07-12 ppdp19/ppdp19.tex
[ ppdp19 ] proof reading for final version

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,screen,fleqn]{acmart} % review for draft
%% DRAFT only:
%\settopmatter{printacmref=false} % Removes citation information below abstract
%\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
\sloppy

\newcommand\hmmax{0}
\newcommand\bmmax{0}
\usepackage{bm} % bold math

@@

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.
The normal form is obtained by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
(\emph{reflecting})
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms.   More precisely, base types are interpreted
as presheaves or sets of neutral term
families, in order to facilitate the generation of
fresh variables of base type during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
\emph{function} type can be reflected into the semantics as functions
applying the variable to their reified argument,
forming a neutral term.  This mechanism provides us with unknowns of
base and function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to \emph{sum} types (logically, disjunctions, and categorically,
weak coproducts), this reflection trick does not work anymore.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
such information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE;  instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the respective injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the leaves
are occupied by non-monadic values \cite{altenkirchUustalu:flops04}.
In this article we only consider weak sum types, producing non-unique
normal forms.  In particular, case trees are not normalized, allowing,
\eg, redundant case splits (\ie, asking the same question twice).
Further, the order of case splits is not normalized: changing the
order in which the questions are asked (commuting case splits),
alters the normal form.
The model would need refinement for extensional sums (strong
coproducts) with unique normal forms
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.

@@

In this work, we formulate NbE for CBPV
(Section~\ref{sec:cbpv}).
% , with the aim to investigate
% later whether CBN and CBV NbE can be recovered from CBPV NbE via the
% standard translations of the CBN and CBV calculi into CBPV.
%
In contrast to the normal forms of CBN NbE, which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models, CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function, and more generally, negative
types.  Non-invertible eliminations are also happening in a chain when
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

In our presentation of the various lambda calculi we ignore the
concrete syntax, only consider the abstract syntax obtained by the
Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
nodes are rule invocations.  Thus, an intrinsically typed,
nameless syntax is most natural,
and our syntactic classes are all presheaves over the category of
typing contexts and renamings.  The use of presheaves then smoothly
extents to the semantic constructions
\cite{catarina:csl93,altenkirchHofmannStreicher:ctcs95}.

@@

Our design choices were guided by an Agda formalization of sections 2
(complete, see \url{https://andreasabel.github.io/ipl/html/NfModelMonad.html})
and 4 (partial, see \url{https://andreasabel.github.io/ipl/html-focusing/Polarized.html}).
Agda was particularly helpful to correctly handle the renamings
abundantly present when working with presheaves.

@@

Simple types shall be distinguished into positive types $P$ and
negative types $N$, depending on their root type former (for now).  Function
($\To$) and product types ($\times$ and $\tyone$) are negative,
while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
\[
\begin{array}{lrl@{\qquad}l}
  A,B,C & ::= & P \mid N & \mbox{simple types} \\
  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
  N     & ::= & \tyone \mid A \times B \mid A \To B
  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
  % N     & ::= & A \To B \mid A \times B \mid \tyone
    & \mbox{negative types} \\
\end{array}
\]
% Simple types shall be given by the grammar
% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
% %   \mid \tyzero$}.
% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
%   \mid \tyzero$}.
% Herein, $a$ ranges over atomic types aka base types.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\quad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\quad
  \nru{\tapp}
    {A \To B \from \Gamma \quad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\tyone \from \Gamma}
\quad
  \nru{\tpair}
    {A_1 \from \Gamma \quad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\quad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\quad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \quad B \from \cext \Gamma A_1
      \quad B \from \cext \Gamma A_2
    }{B \from \Gamma}
\\[2ex]
  \nru{\tabort}
    {\tyzero \from \Gamma}
    {B \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism, where $A \from \Gamma$ is to be read as
``$A$ follows from $\Gamma$''.  Using shorthand
\fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
a term such as
\[
  \tabs\,(\tabs\,(\tpair\,
    \vv1\,
    (\tabs\,(\tapp\,\vv1\,\vv0))))
\]
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to the imagination of the reader,
as we shall not be concerned with considering
concrete terms in this article.

@@

% that does not guarantee any relation between its input term $t$ and
% output normal form $u$.
%
% \emph{Normalization by evaluation} (NbE) decomposes normalization as
% \fbox{\(
%   \norm[\Gamma] A (t : A \from \Gamma)
%   = \down[\Gamma] A \dent t {\idenv\Gamma}
% \)}
% which is
% \emph{evaluation}
% \nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \Den \Gamma \Delta \to
%   \Den A \Delta$}
% in the identity environment \nofbox{$\idenv\Gamma : \Den \Gamma \Gamma$}
% followed by
% \emph{reification}
% \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
\emph{Normalization by evaluation} (NbE)
%\fbox{\(
%  \norm[\Gamma] A (t : A \from \Gamma)
\[
  \tnorm\, (t : A \from \Gamma)
  = \down A \dent t {\idenv\Gamma}
\]
%\)}
decomposes normalization into
\emph{evaluation}
\[%\nofbox{$
\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
  \den A
\]%$}
in the identity environment
\[%\nofbox{$
\idenv\Gamma : \den \Gamma \Gamma
\]%$}
followed by
\emph{reification}
\[%\nofbox{$
\down A : \den A \todot \tNf A
\]%$}
(aka quoting).
The role of evaluation is to produce from a term the corresponding
semantic (\ie, meta-theoretic) function, which is finally reified to a normal form.
Since we are evaluating open terms $t$, we need to supply an environment
$\idenv\Gamma$ which will map the free indices of $t$ to corresponding
\emph{unknowns}.  To accommodate unknowns in the semantics, types $A$
are mapped to presheaves $\den A$ (rather than just sets), and in
particular each base type $\tybase$ is mapped to the presheaf
$\tNe\,\tybase$ with the intention that the neutrals take the role of
the unknowns.  The mapping $\up A : \tNe\,A \todot \den A$ from
neutrals to unknowns is called \emph{reflection} (aka unquoting), and defined mutually
with reification by induction on type $A$.

@@

We are looking for a monad $\C$, called a \emph{cover monad}, that
takes the role of the \emph{oracle} alluded to in the introduction,
and offers us these services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{l@{~}c@{~}l@{\quad}l}
\trunNf[\C] & : & \C\,(\tNf\, P) \todot \tNf\, P
%\\ & & \mbox{run the monad ($\tNf$ only)}
\\
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
%\\ & & \mbox{case on absurd neutral}
\\
\tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
\\
  && {} \to \C\,\B\,(\cext \Gamma {A_1})
  \to \C\,\B\,(\cext \Gamma {A_2})
  \to \C\,\B\,\Gamma
%\\ & & \mbox{case on neutral}
\\
\end{array}
\]
Method $\trunNf[\C]$ enables us to \emph{run} a monadic computation of
a normal form, $\C\,(\tNf\,P)\,\Gamma$, and delivers a normal form
$\Nf P \Gamma$.  However, $\trunNf[\C]$ is not just the inverse of
$\treturn[\C]$; instead, the information contained in $\C$ is reified and
becomes part of the normal form delivered by $\trunNf[\C]$.

@@

Method $\tcase^\C_\Gamma$ allows us, in a computation
$\C\,\B\,\Gamma$,
to ask the oracle about a neutral
term $\NeP{A_1 + A_2}\Gamma$ of sum type.  We have to supply
handlers for both possible answers:
a computation $\C\,\B\,(\cext\Gamma{A_1})$ that can utilize an
additional hypothesis of type $A_1$ in the case of a left injection,
and analogously a computation $\C\,\B\,(\cext\Gamma{A_2})$ for the
case of a right injection.

The similarity of the monad services $\tabort^\C$ and $\tcase^\C$ to
the corresponding constructors for normal forms, or their
generalization of \Cref{rem:abort}, is hard to miss.
Unsurprisingly, just \emph{case trees} are a first instance of a cover monad,
% To make things concrete,
% we shall immediately construct an instance of such a cover monad:
the free cover monad
$\Cov$ defined
%inductively by
as an inductive family with constructors
$\treturn[\Cov]$, $\tabort^{\Cov}$, and
$\tcase^{\Cov}$.
% TODO: Shouldn't this better be trees with nullary and binary nodes
% and leaves always an element of A?
One can visualize an element $c : \Cov\,\A\,\Gamma$ as a
binary case tree whose inner nodes ($\tcase$) are labeled by neutral terms
of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
and $A_2$, resp.  Leaves are either
labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind
(aka Kleisli extension) replaces
these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$,
which flattens a 2-level case tree, \ie, a case tree with case trees as
leaves, into a single one.
Finally $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
and $\tabort^{\Cov}$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

% the following constructions:
% \begin{gather*}
%   \nru{\treturn[\Cov]}{\A\,\Gamma}{\Cov\,\A\,\Gamma}
% \end{gather*}

Using the services of a generic cover monad $\C$, we can complete our semantics:
%\begin{gather*}
\[
\def\arraystretch{1.3}
\begin{array}{lcl}
  \den   \tybase & = & \C\,(\tNe\,\tybase) \\
  \up    \tybase & = & \treturn[\C] \\
  \down  \tybase & = & \trunNf[\C] \comp \tmap[\C]\tne \\[1ex]
% \end{array}
% \qquad
% \begin{array}{lcl}
  \den   \tyzero & = & \C\,\hat0 \\
  \up    \tyzero & = & \tabort^\C \\
  \down  \tyzero & = & \trunNf[\C] \comp \tmap[\C]\tmagic \\[1ex]
% \end{array}
% \]
% \[
% \begin{array}{lcl}
  \den {A+B} & = & \C\,(\den A \hatplus \den B) \\
  \upG {A+B} \; u & = & \tcase^\C\,u
    \,(\treturn[\C]\,(\iota_1\,\tfresh^A_\Gamma)) \\ & & \qquad\quad
    \,(\treturn[\C]\,(\iota_2\,\tfresh^B_\Gamma))
    \\
  \down {A+B} & = & \trunNf[\C] \comp
    \tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B] \\
\end{array}
\]
% \[
% \begin{array}{lcl}
%   \den {A+B} & = & \C\,(\den A \hatplus \den B) \\
%   \upG {A+B} \; u & = & \tcase^\C\,u
%     \,(\treturn[\C]\,(\iota_1\,\tfresh^A_\Gamma))
%     \,(\treturn[\C]\,(\iota_2\,\tfresh^B_\Gamma))
%     \\
%   \downG {A+B} \; m & = & \trunNf[\C]\,
%     (\tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B]\;m) \\
% \end{array}
% \]
%\end{gather*}

>>>>

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,screen,fleqn]{acmart} % review for draft
%% DRAFT only:
%\settopmatter{printacmref=false} % Removes citation information below abstract
%\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in first column
\sloppy
%\usepackage{array}
\newcommand\hmmax{0}
\newcommand\bmmax{0}
\usepackage{bm} % bold math

@@

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter that evaluates closed terms to compute the
normal form of an open term, \ie, a term that may contain free variables.
The normal form is obtained by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
(\emph{reflecting})
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms.   More precisely, base types are interpreted
as presheaves, or sets of neutral term
families, in order to facilitate the generation of
fresh variables of base type during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
\emph{function} type can be reflected into the semantics as functions
applying the variable to their reified argument,
forming a neutral term.  This mechanism provides us with unknowns of
base and function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to \emph{sum} types (logically, disjunctions, and categorically,
weak coproducts), this reflection trick does no longer work.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
such information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE;  instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the respective injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the inner
nodes are labeled with unknowns and the leaves
with non-monadic values \cite{altenkirchUustalu:flops04}.

In this article, we only consider weak sum types, lacking the universal
property of coproducts.  As a consequence, terms with the same
denotation may have different normal forms under NbE.
In particular, case trees are not normalized, allowing,
\eg, redundant case splits (\ie, asking the same question twice).
Further, the order of case splits is not normalized: changing the
order in which the questions are asked (commuting case splits),
alters the normal form.
The model would need refinement for extensional sums (strong
coproducts) with unique normal forms
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17},
\ie, where NbE decides equality of denotations.

@@

In this work, we formulate NbE for CBPV
(Section~\ref{sec:cbpv}).
% , with the aim to investigate
% later whether CBN and CBV NbE can be recovered from CBPV NbE via the
% standard translations of the CBN and CBV calculi into CBPV.
%
In contrast to the normal forms of CBN NbE---which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models---CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function, and more generally, negative
types.  Non-invertible eliminations are also happening in a chain when
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

In our presentation of the various lambda calculi we ignore the
concrete syntax, only considering the abstract syntax obtained by the
Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
nodes are rule invocations.  Thus, an intrinsically typed,
nameless syntax is most natural,
and our syntactic classes are all presheaves over the category of
typing contexts and renamings.  The use of presheaves then smoothly
extents to the semantic constructions
\cite{catarina:csl93,altenkirchHofmannStreicher:ctcs95}.

@@

Our design choices were guided by an Agda formalization of sections 2
(complete, see
\url{https://andreasabel.github.io/ipl/html/NfModelMonad.html}),
3 (in a variantion, see \url{https://andreasabel.github.io/ipl/html-cbpv/NfCBPV.html})
and 4 (partial, see \url{https://andreasabel.github.io/ipl/html-focusing/Polarized.html}).
Agda was particularly helpful to correctly handle the renamings
abundantly present when working with presheaves.

@@

Simple types shall be distinguished into positive types $P$ and
negative types $N$, depending on their root type former (for now).  Function
($\To$) and product types ($\times$ and $\tyone$) are negative,
while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
\[
\begin{array}{lrl@{\qquad}l}
  A,B,C & ::= & P \mid N & \mbox{simple types} \\
  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
  N     & ::= & \tyone \mid A \times B \mid A \To B
  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
  % N     & ::= & A \To B \mid A \times B \mid \tyone
    & \mbox{negative types} \\
\end{array}
\]
% Simple types shall be given by the grammar
% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
% %   \mid \tyzero$}.
% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
%   \mid \tyzero$}.
% Herein, $a$ ranges over atomic types aka base types.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\quad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\quad
  \nru{\tapp}
    {A \To B \from \Gamma \quad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\tyone \from \Gamma}
\quad
  \nru{\tpair}
    {A_1 \from \Gamma \quad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\quad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\quad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \quad B \from \cext \Gamma A_1
      \quad B \from \cext \Gamma A_2
    }{B \from \Gamma}
\\[2ex]
  \nru{\tabort}
    {\tyzero \from \Gamma}
    {B \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism, where $A \from \Gamma$ is to be read as
``$A$ follows from $\Gamma$''.  Using shorthand
\fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
a term such as
\[
  \tabs\,(\tabs\,(\tpair\,
    \vv1\,
    (\tabs\,(\tapp\,\vv1\,\vv0))))
\]
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to the imagination of the reader,
as we shall not be concerned with considering
\emph{concrete} terms in this article.

@@

% that does not guarantee any relation between its input term $t$ and
% output normal form $u$.
%
% \emph{Normalization by evaluation} (NbE) decomposes normalization as
% \fbox{\(
%   \norm[\Gamma] A (t : A \from \Gamma)
%   = \down[\Gamma] A \dent t {\idenv\Gamma}
% \)}
% which is
% \emph{evaluation}
% \nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \Den \Gamma \Delta \to
%   \Den A \Delta$}
% in the identity environment \nofbox{$\idenv\Gamma : \Den \Gamma \Gamma$}
% followed by
% \emph{reification}
% \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
The method
\emph{normalization by evaluation} (NbE)
%\fbox{\(
%  \norm[\Gamma] A (t : A \from \Gamma)
\[
  \tnorm\, (t : A \from \Gamma)
  = \down A \dent t {\idenv\Gamma}
\]
%\)}
decomposes normalization into
\emph{evaluation}
\[%\nofbox{$
\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
  \den A
\]%$}
in the identity environment
\[%\nofbox{$
\idenv\Gamma : \den \Gamma \Gamma
\]%$}
followed by
\emph{reification}
\[%\nofbox{$
\down A : \den A \todot \tNf A
\]%$}
(aka quoting).
The role of evaluation is to produce from a term the corresponding
semantic (\ie, meta-theoretic) function, which is finally reified to a normal form.
Since we are evaluating open terms $t$, we need to supply an environment
$\idenv\Gamma$ which will map the free indices of $t$ to corresponding
\emph{unknowns}.  To accommodate unknowns in the semantics, types $A$
are mapped to presheaves $\den A$ (rather than just sets), and in
particular each base type $\tybase$ is mapped to the presheaf
$\tNe\,\tybase$ with the intention that the neutrals take the role of
the unknowns.  The mapping $\up A : \tNe\,A \todot \den A$ from
neutrals to unknowns is called \emph{reflection} (aka unquoting), and defined mutually
with reification by induction on type $A$.

@@

We are looking for a monad $\C$, called a \emph{cover monad}, that
takes the role of the \emph{oracle} alluded to in the introduction,
and offers us the following services:
% \begin{enumerate}
% \item Elimination of absurd neutrals
% $\tabort^\C : \tNe\, \tyzero \todot \C\,\A$.
% \item Case distinction on neutrals of sum type
% $\tcase^\C_\Gamma : \Ne {(A + B)} \Gamma
%   \to \C\,\A\,(\cext \Gamma A)
%   \to \C\,\A\,(\cext \Gamma B)
%   \to \C\,\A\,\Gamma$.
% \item Run the monad to produce a normal form
% $\trun^\C : \C\,(\Nf A) \todot \Nf A$.
% \end{enumerate}
\[
\begin{array}{l@{~}c@{~}l@{\quad}l}
\trunNf[\C] & : & \C\,(\tNf\, P) \todot \tNf\, P
%\\ & & \mbox{run the monad ($\tNf$ only)}
\\
\tabort^\C & : & \tNe\, \tyzero \todot \C\,\B
%\\ & & \mbox{case on absurd neutral}
\\
\tcase^\C_\Gamma & : & \Ne {(A_1 + A_2)} \Gamma
\\
  && {} \to \C\,\B\,(\cext \Gamma {A_1})
  \to \C\,\B\,(\cext \Gamma {A_2})
  \to \C\,\B\,\Gamma
%\\ & & \mbox{case on neutral}
\\
\end{array}
\]
Method $\trunNf[\C]$ enables us to \emph{run} a monadic computation of
a normal form, $\C\,(\tNf\,P)\,\Gamma$, and delivers a normal form
$\Nf P \Gamma$.
% However, $\trunNf[\C]$ is not just the inverse of
% $\treturn[\C]$; instead, the information contained in $\C$ is reified and
% becomes part of the normal form delivered by $\trunNf[\C]$.
While $\trunNf[\C]$ is a left inverse of
$\treturn[\C]$, the converse is not true.
Instead, the information contained in $\C$ is reified and
becomes part of the normal form delivered by $\trunNf[\C]$.
Technically, $(\tNf\,P, \trunNf[\C])$ is a
\emph{monad algebra} for monad $\C$.
Besides $\trunNf[\C] \circ \treturn[\C] = \tid_{\tNf\,P}$, the square
$\trunNf[\C] \circ \tjoin[\C] = \trunNf[\C] \circ
 \tmap[\C]\trunNf[\C]$ stands.  This means that given a case tree whose
leaves are again case trees ending in normal forms, it does not
matter whether we first join the nested case trees and then assemble
the normal form, or whether we first turn the inner case trees into
normal forms via $\tmap[\C]\trunNf[\C]$, and then the outer one.

@@

Method $\tcase^\C_\Gamma$ allows us, in a computation
$\C\,\B\,\Gamma$,
to ask the oracle about a neutral
term $\NeP{A_1 + A_2}\Gamma$ of sum type.  We have to supply
handlers for both possible answers:
a computation $\C\,\B\,(\cext\Gamma{A_1})$ that can utilize an
additional hypothesis of type $A_1$ in the case of a left injection,
and analogously a computation $\C\,\B\,(\cext\Gamma{A_2})$ for the
case of a right injection.
\newcommand{\figdenpos}{
\begin{figure}[bthp]
\hrulefill
\[
\def\arraystretch{1.3}
\begin{array}{lcl}
  \den   \tybase & = & \C\,(\tNe\,\tybase) \\
  \up    \tybase & = & \treturn[\C] \\
  \down  \tybase & = & \trunNf[\C] \comp \tmap[\C]\tne \\[1ex]
% \end{array}
% \qquad
% \begin{array}{lcl}
  \den   \tyzero & = & \C\,\hat0 \\
  \up    \tyzero & = & \tabort^\C \\
  \down  \tyzero & = & \trunNf[\C] \comp \tmap[\C]\tmagic \\[1ex]
% \end{array}
% \]
% \[
% \begin{array}{lcl}
  \den {A+B} & = & \C\,(\den A \hatplus \den B) \\
  \upG {A+B} \; u & = & \tcase^\C\,u
    \,(\treturn[\C]\,(\iota_1\,\tfresh^A_\Gamma)) \\ & & \qquad\quad
    \,(\treturn[\C]\,(\iota_2\,\tfresh^B_\Gamma))
    \\
  \down {A+B} & = & \trunNf[\C] \comp
    \tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B] \\
\end{array}
\]
\hrulefill
  \caption{Interpretation of positive types.}
  \label{fig:denpos}
\end{figure}
% \[
% \begin{array}{lcl}
%   \den {A+B} & = & \C\,(\den A \hatplus \den B) \\
%   \upG {A+B} \; u & = & \tcase^\C\,u
%     \,(\treturn[\C]\,(\iota_1\,\tfresh^A_\Gamma))
%     \,(\treturn[\C]\,(\iota_2\,\tfresh^B_\Gamma))
%     \\
%   \downG {A+B} \; m & = & \trunNf[\C]\,
%     (\tmap[\C] [\tinj_1 \comp \down A,\; \tinj_2 \comp \down B]\;m) \\
% \end{array}
% \]
} % END \newcommand
\figdenpos

The similarity of the monad services $\tabort^\C$ and $\tcase^\C$ to
the corresponding constructors for normal forms, or their
generalization of \Cref{rem:abort}, is hard to miss.
Unsurprisingly, just \emph{case trees} are a first instance of a cover monad:
the free cover monad
$\Cov$ defined
as an inductive family with constructors
$\treturn[\Cov]$, $\tabort^{\Cov}$, and
$\tcase^{\Cov}$.
% TODO: Shouldn't this better be trees with nullary and binary nodes
% and leaves always an element of A?
One can visualize an element $c : \Cov\,\A\,\Gamma$ as a
binary case tree whose inner nodes ($\tcase$) are labeled by neutral terms
of sum type $A_1+A_2$ and its two branches by the context extensions $A_1$
and $A_2$, resp.  Leaves are either
labeled by a neutral term of empty type $\tyzero$ (see $\tabort$),
or by an element of $\A$ (see $\treturn$).  Functoriality amounts to
replacing the labels of the $\treturn$-leaves, and the monadic bind
(aka Kleisli extension) replaces
these leaves by further case trees.  Bind is realized via $\tjoin[\Cov]$,
which flattens a 2-level case tree, \ie, a case tree with case trees as
leaves, into a single one.
Finally, $\trunNf[\Cov]$ is a simple recursion on the tree, replacing $\tcase^{\Cov}$
and $\tabort^{\Cov}$
by the $\tcase$ and $\tabort$ constructions on normal forms, and
$\treturn[\Cov]$ by the identity.

Using the services of a generic cover monad $\C$, we can complete our
semantics, see \Cref{fig:denpos}.
